Which fifteen-minutes neighborhoods are dead-ends? An analysis of the network attributes of fifteen-minute pedsheds

### 2024-10-21 paper/paper.Rmd
Training of evolutionary trees complete; saved trained trees to ensure that results are consistent for paper.

```{r load-packages}
library(broom) # Convert Statistical Objects into Tidy Tibbles
library(cowplot)
library(dplyr) # A Grammar of Data Manipulation
library(evtree) # Evolutionary Learning of Globally Optimal Trees
library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(glue) # Interpreted String Literals
library(gt) # Easily Create Presentation-Ready Display Tables
library(rpart) # Recursive Partitioning and Regression Trees
library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
library(patchwork) # The Composer of Plots
library(partykit)
library(tidyr) # Tidy Messy Data
library(tree) # Classification and Regression Trees
library(sf) # Simple Features for R
library(skimr) # Compact and Flexible Summaries of Data
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(units) # Measurement Units for R Vectors
library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

@@

# Analysis-ready data. See notebooks in folder data-raw for details
load(file = glue::glue(here::here(),
                       "/data/amenities_walkshed_da.rda"))
load(file = glue::glue(here::here(),
                       "/data/hamilton_net.rda"))
load(file = glue::glue(here::here(),
                       "/data/urban_hammer_da.rda"))
load(file = glue::glue(here::here(),
                       "/data/walksheds_da.rda"))
load(file = glue::glue(here::here(),
                       "/data/walksheds_da_net_vars.rda"))
```

The aim is to train a classification tree to profile urban and suburban walksheds based on the attributes of the network. We need to drop 1 observation that has an NA in `transitivity`:

```{r prepare-data-frame-for-analysis}
# Use normalized motifs because the number of motifs will depend on the number of edges. Divide by edges to obtain a comparable number
# Remove girth which is problematic for modeling
d <- walksheds_da_net_vars |>
  select(-girth) |>
  drop_na() |>
  mutate(`Normalized Motifs (3)` = motifs_3/n_edges,
         `Normalized Motifs (4)` = motifs_4/n_edges) |>
  rename(`Edge Density` = edge_density)
```

```{r train-evolutionary-tree}
#seed1 <- 908544
set.seed(908544)
ev.tree <- evtree::evtree(Type ~ ., 
                data = d |> select(-c(GeoUID, 
                                      motifs_3, 
                                      motifs_4, 
                                      n_edges, 
                                      n_nodes)))
```

```{r create-tree-plot}
party_plot <- ev.tree |>
  ggparty() +
  geom_edge() +
  #geom_edge_label(size = 2.5) +
  geom_node_label(aes(label = splitvar ),
                  id = "inner",
                  size = 2, nudge_y = -0.022) +
  geom_node_splitvar(id = "inner", size = 4) +
  geom_edge_label(mapping = aes(label = prettyNum(breaks_label, digits = 2)), size = 2.5) +
  # # pass list to gglist containing all ggplot components we want to plot for each
  # # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = Type),
                                        position = position_fill()),
                               #xlab(""),
                               #ylab(""),
                               theme_void(),
                               scale_fill_manual(name = "Urban fabric type",
                    values = c("#FDAE61", "#1A9850")),
                              theme(axis.text = element_blank())))
```

```{r legend-for-tree-plot}
#creating a legend
types_legend <- ggplot(d) +
  geom_bar(aes(x = Type, 
               fill = Type)) +
  scale_fill_manual(name = "Urban fabric type",
                    values = c("#FDAE61", "#1A9850"),
              labels = c("Suburban", 
                         "Urban") +
                theme(legend.position = "bottom")) 
types_legend <- cowplot::get_legend(types_legend)
```

```{r tree-plot, fig.cap="\\label{fig:classification-tree}Classification tree of walksheds by type of urban fabric"}
cowplot::plot_grid(party_plot, legd_renamed, 
                   nrow = 2,
                   align = "v", axis = "l", rel_heights = c(6, 1), rel_widths = c(1, 1))
```

>>>>

```{r load-packages}
library(broom) # Convert Statistical Objects into Tidy Tibbles
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
library(dplyr) # A Grammar of Data Manipulation
library(evtree) # Evolutionary Learning of Globally Optimal Trees
library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(glue) # Interpreted String Literals
library(gt) # Easily Create Presentation-Ready Display Tables
#library(rpart) # Recursive Partitioning and Regression Trees
#library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
#library(patchwork) # The Composer of Plots
#library(partykit) # A Toolkit for Recursive Partytioning
library(tidyr) # Tidy Messy Data
#library(tree) # Classification and Regression Trees
library(sf) # Simple Features for R
library(skimr) # Compact and Flexible Summaries of Data
#library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(units) # Measurement Units for R Vectors
library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

@@

# Load trained evolutionary trees
load(file = glue::glue(here::here(),
                       "/data/trees.rda"))

# Analysis-ready data. See notebooks in folder data-raw for details
load(file = glue::glue(here::here(),
                       "/data/amenities_walkshed_da.rda"))
load(file = glue::glue(here::here(),
                       "/data/hamilton_net.rda"))
load(file = glue::glue(here::here(),
                       "/data/urban_hammer_da.rda"))
load(file = glue::glue(here::here(),
                       "/data/walksheds_da.rda"))
load(file = glue::glue(here::here(),
                       "/data/walksheds_da_net_vars.rda"))
load(file = glue::glue(here::here(),
                       "/data/walksheds_validation.rda"))
load(file = glue::glue(here::here(),
                       "/data/walksheds_validation_net_vars.rda"))
```

```{r}
# Bind the centroid and non-centroid walksheds:
walksheds <- bind_rows(walksheds_da,
                       walksheds_validation) |>
  mutate(origin_type = rep(c("centroid", "non-centroid"), each = nrow(walksheds_da)))
```

```{r }
# Obtain training and validation datasets by sampling from centroid and non-centroid walksheds:

# Set seed for replicability
seed <- 3626425
set.seed(seed)

# slice_sample() will select one of two rows for each DA at random
walksheds_train <- walksheds |>
  group_by(GeoUID) |>
  slice_sample(prop = 0.5)

# The antijoin will keep the rows in walksheds that do not appear in walksheds_train
walksheds_test <- walksheds |>
  anti_join(walksheds_train |>
              st_drop_geometry() |>
              select(GeoUID, origin_type),
            by = c("GeoUID", "origin_type"))
```

```{r bind-centroid-non-centroid-network-variables}
# Bind the centroid and non-centroid dataframes with walkshed network variables:
walksheds_net_vars <- bind_rows(walksheds_da_net_vars,
                                walksheds_validation_net_vars) |>
  mutate(origin_type = rep(c("centroid",
                             "non-centroid"), 
                           each = nrow(walksheds_da)))
```

```{r obtain-training-testing-network-variables}
# Now filter from the network variables to obtain training and test datasets:
walksheds_train_net_vars <- walksheds_net_vars |>
  semi_join(walksheds_train |>
              st_drop_geometry() |>
              select(GeoUID, origin_type),
            by = c("GeoUID", "origin_type"))

walksheds_test_net_vars <- walksheds_net_vars |>
  semi_join(walksheds_test |>
              st_drop_geometry() |>
              select(GeoUID, origin_type),
            by = c("GeoUID", "origin_type"))
```

```{r training-and-testing-dataframes-for-evolutionary-tree}
d <- walksheds_train_net_vars |>
  select(-girth) |>
  drop_na() |>
  mutate(normalized_motifs_3 = motifs_3/n_edges,
         normalized_motifs_4 = motifs_4/n_edges)

d_test <- walksheds_test_net_vars |>
  select(-girth) |>
  drop_na() |>
  mutate(normalized_motifs_3 = motifs_3/n_edges,
         normalized_motifs_4 = motifs_4/n_edges)
```

```{r best-tree}
# Retrieve the best evolutionary tree. The algorithm of evolutionary trees has some randomness due to random initial values. We compared a number of trees and chose the one that gave the best fit in terms of missclassification and size of the model. See notebook `05-Walkshed-Profiles.Rmd in folder `data-raw` for details.

best_evtree <- trees[[4]]
```

```{r generate-plot-best-tree}
# Round the labels of the edges. The relevant information is in a dataframe produced by ggparty()
ggparty_plot <- best_evtree |>
  ggparty()

# Round the number of digits and keep the inequality signs
ggparty_data <- ggparty_plot$data |>
  mutate(breaks_label = paste0(substr(breaks_label, 4, 5), 
                               substr(breaks_label,10, 15)),
         splitvar = case_when(splitvar == "edge_density" ~ "Edge Density",
                              splitvar == "transitivity" ~ "Transitivity",
                              splitvar == "normalized_motifs_3" ~ "Normalized Motifs (3)",
                              is.na(splitvar) ~ NA))

# Replace the dataframe in the ggplot object
ggparty_plot$data <- ggparty_data

evtree_plot <- ggparty_plot +
  geom_edge() +
  geom_node_splitvar(id = "inner", size = 3) +
  # parse should be set to FALSE so the labels do not try to parse the inequality signs
  geom_edge_label(mapping = aes(label = breaks_label),
                  parse = FALSE,
                  size = 2.5) +
  # # pass list to gglist containing all ggplot components we want to plot for each
  # # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = Type),
                                        position = position_fill()),
                               xlab(""),
                               ylab(""),
                               scale_fill_manual(name = "Urban fabric type",
                                                 values = c("#FDAE61", "#1A9850")),
                               theme(axis.text.y = element_blank(),
                                     panel.background = element_rect(fill = NA),
                                     axis.ticks = element_blank(),
                                     plot.background =  element_rect(fill = NA)))) +
  geom_node_label(line_list = list(aes(label = paste0("Leaf ", ifelse(id >= 6, id - 3, id - 2)))),
                  line_gpar = list(list(size = 7, parse = F)),
                  ids = "terminal",
                  nudge_x = 0.02,
                  nudge_y = -0.4)
```

```{r plot-evtree, fig.cap="\\label{fig:evolutionary-tree}Results of training evolutionary tree on the attributes of the network. The dependent variable is Urban Fabric, a binary variable with values "Urban" (green) and "Suburban" (orange)"}
evtree_plot

ggsave(evtree_plot,
       file = glue::glue(here::here(),"/plots/evtree.png"),
       width = 7,
       units = "in")
```
